---
title: "FML - Clustering - April1st"
author: "Aamir Sohaib"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(flexclust)
library(factoextra)
```


# Loading dataset into R environment 

```{r}
pharma <- read.csv("/Users/roc/Downloads/Pharmaceuticals-1.csv")
```

# Understanding basics of finacial data 

```{r}
str(pharma)
summary(pharma)
```

# PART (a):Cluster using only numerical variables

```{r}
df_num <- pharma[, 3:11]
rownames(df_num) <- pharma$Name
print(df_num)
```


# Z-Score Normalization

# Now I scale the data using z-score normalization to remove scale bias. Without scaling, variables with larger ranges would dominate the distance calculations unfairly.

```{r}
df_scaled <- scale(df_num)
print(df_scaled)
```

# Now I compute Euclidean distances between all pairs of firms

```{r}
distance <- get_dist(df_scaled)
```

```{r}
fviz_dist(distance,gradient= list(low="#00AFBB",mid = "white", high= "#FC4E07"),
          lab_size = 6) +labs(title ="Distance Matrix: Pharmaceutical Firms")
```

# Now I determine Optimal Number of Clusters (k)

# Method-1: Elbow Method

```{r}
set.seed(123)
fviz_nbclust(df_scaled,kmeans, method = "wss") +labs(title= "Elbow Method: Optimal Number of Clusters",x ="Number of Clusters (k)",y = "Total Within-Cluster Sum of Squares (WSS)")
```

# Method 2: Silhouette Method

```{r}
set.seed(123)
fviz_nbclust(df_scaled,kmeans, method = "silhouette") +labs(title ="Silhouette Method: Optimal Number of Clusters",x ="Number of Clusters (k)",y ="Average Silhouette Width")
```

# Now I run K-Means clustering with k=5. Based on Elbow and Silhouette methods, k = 5 is selected as optimal.

```{r}
set.seed(123)
k5 <- kmeans(df_scaled,centers =5,nstart= 25)
print(k5$centers)
```

# to view size of each cluster

```{r}
print(k5$size)
```

# to view cluster assignment for each firm

```{r}
print(k5$cluster)
```
# Now I visualize the clusters

```{r}
fviz_cluster(k5, data=df_scaled,palette = c("#2E9FDF","#E7B800","#FC4E07","#00BA38","#9b59b6"), ellipse.type= "euclid",repel=TRUE,ggtheme = theme_minimal()) +
  labs(title ="K-Means Clustering (k=5): Pharmaceutical Firms")
```

# PART (b): Interpretetion of clusters based on numerical variables

```{r}
pharma$Cluster <- k5$cluster
cluster_summary <- pharma %>%
  group_by(Cluster) %>%
  summarise(
    Firms        = n(),
    Avg_MarketCap =mean(Market_Cap),
    Avg_Beta      =mean(Beta),
    Avg_PE_Ratio  =mean(PE_Ratio),
    Avg_ROE       =mean(ROE),
    Avg_ROA       =mean(ROA),
    Avg_AssetTO   =mean(Asset_Turnover),
    Avg_Leverage  =mean(Leverage),
    Avg_RevGrowth =mean(Rev_Growth),
    Avg_NetMargin =mean(Net_Profit_Margin))
print(cluster_summary)
```

```{r}
pharma %>%
  select(Name, Cluster) %>%
  arrange(Cluster) %>%
  print()
```

# PART (c): Pattern in non-numerical variables(10-12)

```{r}
cat("\n Median Recommendation by Cluster \n")
pharma %>%
  select(Name, Cluster, Median_Recommendation) %>%
  arrange(Cluster) %>%
  print()
```

# Now I count each recommendation type per cluster

```{r}
rec_pattern <- pharma %>%
  group_by(Cluster, Median_Recommendation) %>%
  summarise(Count = n(), .groups = "drop")
print(rec_pattern)
```

# Pattern: Location (HQ Country) per Cluster

```{r}
cat("\n Location (HQ Country) by Cluster \n")
loc_pattern <- pharma %>%
  group_by(Cluster, Location) %>%
  summarise(Count = n(), .groups = "drop")
print(loc_pattern)
```

# Pattern: Stock Exchange per Cluster

```{r}
cat("\n Stock Exchange by Cluster \n")
exc_pattern <- pharma %>%
  group_by(Cluster, Exchange) %>%
  summarise(Count = n(), .groups = "drop")
print(exc_pattern)
```

# PART (d): Name Each Cluster
# Based on cluster_summary above, I assign descriptive names.These names reflect dominant financial characteristics of each group.

```{r}
cluster_names <- data.frame(
  Cluster = 1:5,
  Cluster_Name = c(
    "Large-Cap Market Leaders",
    "High-Risk Speculative Firms",
    "Undervalued Growth Firms",
    "Profitable Dividend Plays",
    "Small Emerging Challengers"))
```

# Merge cluster names into the main dataset

```{r}
pharma_final <- pharma %>%
  left_join(cluster_names, by = "Cluster")
```

# Final output: each firm with its cluster number and descriptive name

```{r}
pharma_final %>%
  select(Name, Cluster, Cluster_Name, Median_Recommendation, Location, Exchange) %>%
  arrange(Cluster) %>%
  print()
```

# Brief Summary
The clustering analysis identified five financially distinct categories within pharmaceutical sector. Large-cap companies i.e. Pfizer, J&J, Merck exhibited superior ROE, ROA, and net profit margins, indicating high operational efficiency.

High leverage companies faced increased financial risk due to lower asset turnover. Firms focused on growth demonstrated robust revenue increases but had thinner profit margins, which suggests aggressive reinvestment strategies.Stable mid cap companies maintained a balance of moderate profitability and low beta, implying reduced market risk. Lastly, small-cap firms displayed elevated PE ratios despite yielding modest returns, reflecting market optimism.

In short, firm size and profitability emerged as most significant financial differentiators among five identified clusters.

Five clusters exhibited unique financial characteristics. Companies such as Pfizer,Johnson & Johnson, and Merck were categorized together as large cap, high profitability leaders. Second cluster included high risk, high leverage firms with fluctuating returns. Third cluster represented undervalued firms that demonstrated robust revenue growth but had moderate profit margins. The fourth group comprised stable, high-margin firms that maintained consistent returns on equity. Fifth cluster encompassed small emerging firms with significant growth potential but lower current profitability.

In terms of non-numerical variables, NYSE-listed US firms were prevalent in most clusters, while "Hold" and "Moderate Buy" were the most common broker recommendations across groups suggesting general market confidence in sector.
